### 2025-07-21 decision-making.qmd
20250721 - Vegas lines and the betting market; delay discounting curves

Aggregation of predictions from multiple people leverages several important features, including cognitive diversity and error cancellation.

Cognitive diversity refers to the representation of individuals with different perspectives because of their "differences in knowledge, training, experience, or thinking styles" [@Rader2017, p. 8].
Cognitive diversity is important because judgments from a cognitively homogeneous group will tend to err *systematically*.
That is, they tend to err in the same direction—either consistently above or below the truth; thus, their errors are correlated.
By contrast, a cognitively diverse group will not tend to err systematically.
Individuals of a cognitively diverse group will bring different areas of expertise (e.g., determination of player skill, player opportunity, matchup strength, etc.) to bear in making their predictions and will thus make different mistakes [@Larrick2024].
Consequently, for the people comprising a cognitively diverse group, their judgments will tend to err *randomly*—where some people's predictions fall above the truth and some people's predictions fall below the truth—i.e., individual judgments "bracket" the truth [@Mannes2014].
Thus, judgments from a cognitively diverse group tend to have uncorrelated errors.

@@

[overconfidence; confirmation bias; variable ratio reinforcement; addiction/delay discounting; wisdom of the crowd, and the challenge of beating "the market"]

@@

@@

Intermittent reinforcement (also called partial reinforcement) means rewarding the behavior sometimes but not everytime.
There are various approaches to intermittent reinforcement, called schedules of reinforcement.
The four primary reinforcement schedules are: fixed interval, fixed ratio, variable interval, adn variable ratio.
The reinforcement schedules are depicted in @fig-reinforcementSchedules, as adapted from @Spielman2020.

>>>>

Aggregation of predictions from multiple people leverages several important features, including cognitive diversity and error cancellation.
Cognitive diversity refers to the representation of individuals with different perspectives because of their "differences in knowledge, training, experience, or thinking styles" [@Rader2017, p. 8].
Cognitive diversity is important because judgments from a cognitively homogeneous group will tend to err *systematically*.
That is, they tend to err in the same direction—either consistently above or below the truth; thus, their errors are correlated.
By contrast, a cognitively diverse group will not tend to err systematically.
Individuals of a cognitively diverse group will bring different areas of expertise (e.g., determination of player skill, player opportunity, matchup strength, etc.) to bear in making their predictions and will thus make different mistakes [@Larrick2024].
Consequently, for the people comprising a cognitively diverse group, their judgments will tend to err *randomly*—where some people's predictions fall above the truth and some people's predictions fall below the truth—i.e., individual judgments "bracket" the truth [@Mannes2014].
Thus, judgments from a cognitively diverse group tend to have uncorrelated errors.

@@

@@

People may make a sports bet because they believe strongly in their predicted outcome and that they "cannot be wrong".
However, the initial Vegas lines represent the aggregation of a massive amount of information (by professional oddsmakers), including team power rankings, injuries, weather, home/away, coaching, matchups, historical betting behavior, advanced statistics, and proprietary algorithms.
Leading up to the game, the Vegas lines are adjusted to balance the total money on each side of the bet (so that the sportsbook can be ensured it makes money).
In this way, the Vegas lines are based on the bets of many, many people and many dollars and reflect market wisdom.
That is, by making a sports bet, you are implicitly claiming that you have information, insight, or a model that makes you better than the market consensus at predicting the outcome—even though half of the dollars bet will lose.
Betting markets, similar to the stock market (as described in @sec-wisdomOfTheCrowd), are often highly efficient and beating them consistently is extremely challenging.
Moreover, people tend to show [confirmation bias](#sec-cognitiveBiasesConfirmation), such that they tend to remember their predictive successes and to forget their failures.
So, when they they get a bet correct, they may be more likely to bet again in the future, especially because sports betting involves [intermittent reinforcement](#sec-reinforcementIntermittent) (in particular, variable ratio reinforcement), which can make it highly addictive.

However, not all people are equally prone to gambling addiction.
One of the most robust risk factors of gambling addiction is a steep delay discounting curve [@Amlung2017; @Weinsztok2021].
The delay discounting curve can be generated from asking respondents a series of hypothetical choices, such as:
1) Would you prefer $10 now or $15 in 1 hour?
2) Would you prefer $10 now or $20 tomorrow?
3) Would you prefer $10 now or $50 in one month?
Based on these questions, one can estimate how much a respondent's valuation of a reward (in this case, money) decreases with the passage of time.
An example of delay discounting curves of two people is in @fig-delayDiscountingCurves.
Some individuals have a shallow delay discounting curve and will prefer more money even if they have to wait, a form of delayed gratification.
Other individuals have a steep delay discounting curve and will prefer obtaining the money now, even if that means gaining less money in the long run, a form of impulsivity.
In particular, individuals who have a steep delay discounting curve are most likely to develop gambling addictions, possibly because they are more sensitive to immediate rewards and are less driven by long-term consequences.

::: {#fig-delayDiscountingCurves}
![](images/delayDiscountingCurves.jpg){width=100% fig-alt="Prototypic hyperbolic delayed reward discounting curves reflecting the discounted subjective value of $100 delayed from 1 day to 1 year The curves reflect the points at which the smaller immediate reward is equal in value to the $100 delayed reward. For example, at a delay of 100 days, $100 has lost ∼50% of its nominal value for the low impulsivity profile and ∼90% of its nominal value for the high impulsivity profile. (Figure and caption adapted from Gray, J. C., & MacKillop, J. (2015). Impulsive delayed reward discounting as a genetically-influenced target for drug abuse prevention: A critical evaluation. *Frontiers in Psychology*, 6. https://doi.org/10.3389/fpsyg.2015.01104)"}

Prototypic hyperbolic delayed reward discounting curves reflecting the discounted subjective value of $100 delayed from 1 day to 1 year The curves reflect the points at which the smaller immediate reward is equal in value to the $100 delayed reward. For example, at a delay of 100 days, $100 has lost ∼50% of its nominal value for the low impulsivity profile and ∼90% of its nominal value for the high impulsivity profile. (Figure and caption adapted from @Gray2015; Gray, J. C., & MacKillop, J. (2015). Impulsive delayed reward discounting as a genetically-influenced target for drug abuse prevention: A critical evaluation. *Frontiers in Psychology*, 6. [https://doi.org/10.3389/fpsyg.2015.01104](https://doi.org/10.3389/fpsyg.2015.01104)).
:::

@@

Intermittent reinforcement (also called partial reinforcement) means rewarding the behavior sometimes but not everytime.
There are various approaches to intermittent reinforcement, called schedules of reinforcement.
The four primary reinforcement schedules are: fixed interval, fixed ratio, variable interval, and variable ratio.
The reinforcement schedules are depicted in @fig-reinforcementSchedules, as adapted from @Spielman2020.
